---
title: "Exercises 6-10"
author: 
- name: Tran Thai Hung
  affiliations:
    - OUCRU biostat
date: 'last-modified'
date-format: '[Last version:] MMMM DD, YYYY'
format: 
  pdf:
    fontsize: 12pt
  html:
    scrollable: true
    code-tools:
      source: true
      toggle: false
      caption: Code    
toc: TRUE
lof: FALSE
lot: FALSE
number-sections: FALSE
thanks: "Please do not circulate."
editor: visual
execute:
  echo: false
  eval: false
---

## RStudio

We will work with the dataset that gives the survival status of passengers on the
Titanic. See [this link](https://hbiostat.org/data/repo/ctitanic3) for a description
of the variables. We provide the dataset in **Excel** format
([Titanic3.xlsx](https://github.com/OUCRU-Modelling/R-training-2025/blob/main/slides/data/Titanic3.xlsx)). It
has three additional columns: `dob` (date of birth), `family` (whether the person
tarvelled with family) and `agecat` (age categorized in age ranges).

We assume that you have created a folder on your computer for the R course. Start the
RStudio program from within that folder. 

You are asked to write the code in an R Script file.  Open a new R Script file via
the menu:

**File → New File → R Script…**

Check whether your working directory is correct. If not, set the working directory
via 

**Session → Set Working Directory → Choose Directory…**


It is useful to learn a couple of RStudio shortcuts that you frequently use. You can
find an overview of them under the **Help  → Keyboard Shortcuts help**. More
shortcuts and tips can be found on the [Appsilon](https://www.appsilon.com/post/rstudio-shortcuts-and-tips) website.

<details>

<summary>[Some useful Keyboard Shortcuts in RStudio for Windows]{style="color:blue"}</summary>

::: callout-note
- **Alt + -**: Insert assignment `<-`
- **F1**: Show function help page when cursor is on function name
- **Ctrl + 1 / Ctrl + 2**: Move focus to Source/Console
- **Tab / Ctrl + Space**: Code suggestion/completion
- **In source editor**:
  - **Ctrl + Enter**: Run current line
  - **Ctrl + Tab**: Switch between tabs in Source Editor

-   **In Console**:
    -   Arrows **Up/Down** to navigate earlier commands
    -   **Ctrl + Up** pops up a listing of latest commands
:::

</details>

On the Mac... (**Hung**)

### Exercise 3: importing data from Excel and first check

Import the dataset by clicking on **Import Dataset** from the **Environment**
pane. Browse for the Excel file. Assign the R object name `titanic` to the dataset
under **Import Options**. Copy the code under **Code Preview:** to your script
file. Click on the "Import" button.

Inspect the data set in spreadsheet-like format that shows up after importing (or
click on the `titanic` object in the **Environment** pane). Use the function `dim` to
find the number of rows (passengers) and columns (variables) in the data set. Issue
the commands
```{r}
#| echo: true
head(titanic,4)
tail(titanic,4)
```
What do you observe? The `readxl` package that is used to import the data stores the
data as a `tibble`, which is a special type of format to store ata in R. Override the
`tibble` format using the function `as.data.frame`. 
```{r}
#| echo: true
titanic <- as.data.frame(titanic)
```
Run the `head` and `tail` functions again. 

Use the `str` function to obtain a summary of the basic characteristics per
variable. Have a look at the columns for `survived` and `age`. Do they have the mode
that you expect?  (You can also use the `mode` function to check their modes.)


### Exercise 4: some columns summarized
A problem with the mode for age is that missing values are coded as a character value
`"NA"` in our Excel file; it is not recognized as an R type NA value. In the
future, you may encounter other representations of missing values, such as `missing`,
`404`, `-99` or `unknown`, which should also be treated as `NA`. Read in the data set
as `titanic` object again, but specify `NA` as the missing value indicator under
**Code Preview**. Your code under **Code Preview** should now look like

```{r}
#| echo: true
library(readxl)
titanic <- read_excel("W:/Teaching/RCourse2023/Exercises/Titanic3.xlsx", 
    na = "NA")
View(titanic)
```

Summarize the whole dataset using the `summary` function. What do you observe for
each of the variables? 

Compute the 5%, 25%, 50%, 75% and 95% quantiles, the inter-quartile range and the
standard deviation for age and fare. You may have to take a look at the `help` files
for the functions `quantile`, `IQR` and `sd`.

For categorical variables, `summary` is not very informative. Use the `table`
function to summarize the variables `pclass`, `survived` and `sex`. Also give a
two-by-two table for sex and survival status using the same `table` function.

```{r}
## survival status
table(titanic$survived)
table(titanic$sex, titanic$survived)
table(titanic$sex, titanic$survived, useNA="always") # has no added value
addmargins(table(titanic$sex, titanic$survived))
```

