---
title: "Exercises Day 3"
author:
- name: Ronald Geskus
  affiliations:
    - OUCRU biostat
- name: Tran Thai Hung
  affiliations:
    - OUCRU biostat
date: 'last-modified'
date-format: '[Last version:] MMMM DD, YYYY'
format: 
  pdf:
    fontsize: 12pt
toc: TRUE
lof: FALSE
lot: FALSE
number-sections: FALSE
thanks: "Please do not circulate."
abstract: ""
execute:
  echo: true
  eval: true
  warning: false
---

## Instructions

Import the data set and load the **ggplot2** package. Change the path to the Excel
file if it is in another folder on your computer. 

```{.r}
#| eval: false
library(readxl)
library(ggplot2)
titanic <- read_excel("data/raw/Titanic3.xlsx", na="NA")
```

```{r}
#| echo: false
library(readxl)
library(ggplot2)
titanic <- read_excel("../slides/data/Titanic3.xlsx", na="NA")
```

## Exercise 1

Plot the price that was paid (variable `fare`) as a function of age. Assign/map the
appropriate variable to the x-axis and y-axis. Next, define the geometric
object. Change the label of fare into "fare paid (GBP)".

```{r warning=TRUE}
ggplot(titanic, aes(age, fare)) + geom_point() + ylab("fare paid (GBP)")
ggplot(titanic, aes(age, fare)) + geom_point() + 
  ylab("fare paid (GBP)") + coord_polar()
```

See what happens if you change the coordinate system to polar.

## Exercise 2
Return to the cartesian coordinate system. Add a second layer that fits the linear
regression line with 95% confidence intervals through the scatterplot (this is seen
as smoothing, choose as method `lm`). Change the overall theme, trying `theme_bw` and
`theme_minimal`. Save the graph as a png file.

```{r}
ggplot(titanic, aes(age, fare)) + geom_point() + 
  ylab("fare paid (GBP)") + geom_smooth(method="lm") + theme_minimal()
ggplot(titanic, aes(age, fare)) + geom_point() + 
  ylab("fare paid (GBP)") + geom_smooth(method="lm") + theme_bw()
ggsave("SavedPlot.png")
```

## Exercise 3
The distribution of fare is very skewed. Therefore, it may be better to plot fare on
a logarithmic scale. Repeat the exercise but now use a log scale along the fare
axis. If you compare the fitted linear trends, you will notice that they do not
coincide. This is for two reasons. First, there were persons that didn't pay any
money. They are removed if the logarithm is taken (you cannot take the log of
zero). Second, the regression line is computed after the scale transformation, hence
based on very different values. 

```{r warning=TRUE}
ggplot(titanic, aes(age, fare)) + geom_point() +
  geom_smooth(method="lm") + theme_bw() + 
  scale_y_log10("fare paid (GBP)")
```

## Exercise 4
We continue with the graph from the last exercise that uses the log scale for
fare. Color the individual data by passenger class. Split the calculation and plot of
the linear regression line by these three groups. 

```{r}
ggplot(titanic, aes(age, fare, color=pclass)) + geom_point() + 
  geom_smooth(method="lm") + theme_bw() + 
  scale_y_log10("fare paid (GBP)")
```

## Exercise 5
Split the graph into two separate ones by sex, next to each other. Add a rug plot to
both sides.

```{r}
ggplot(titanic, aes(age, fare, color=pclass)) + geom_point() +
  geom_smooth(method="lm") + theme_bw() +
  scale_y_log10("fare paid (GBP)") + facet_wrap(~sex) + geom_rug()
```

## Exercise 6
Make a boxplot of age (along y-axis) for each of the passenger classes (along x-axis). See what happens if you flip the coordinates.

```{r}
ggplot(titanic, aes(pclass, age)) + geom_boxplot() +
  xlab("passenger class")
ggplot(titanic, aes(pclass, age)) + geom_boxplot() +
  xlab("passenger class") + coord_flip()
```

## Exercise 7

Add the individual data points to the boxplot. What do you
 observe? Change the default theme if you think it appropriate. 
 
```{r}
ggplot(titanic, aes(pclass, age)) + geom_boxplot() + 
  xlab("passenger class") + geom_point(color="red",alpha=0.2) + 
  theme_minimal()
```

Use the jitter instead of the point geometric object to be able to discriminate them,
give the points a red color and make them transparent by setting the transparency
parameter (alpha) to 0.2. 

```{r}
ggplot(titanic, aes(pclass, age)) + geom_boxplot() +
  xlab("passenger class") + geom_jitter(color="red", alpha=0.2) + 
  theme_minimal()
```

Replace the boxplot by a violin plot.

```{r}
ggplot(titanic, aes(pclass, age)) + geom_violin() + 
  xlab("passenger class") + geom_jitter() + theme_minimal()
```


## Exercise 8
We make a new graph in which we perform an exploratory analysis of the relation between survival and age, sex and passenger class. Define the assignment/mapping to the x-axis (`age`) and y-axis (`survived`). Make a scatterplot of survival status as a function of age.

```{r}
ggplot(titanic, aes(age,survived)) + geom_point() 
```


In order to better see the individual outcomes, use jitter with a height parameter of
0.05 and a width parameter of 0 (why would you set width at 0?); set transparency alpha to 0.5. Add a smooth layer using the **loess** method with 95% confidence intervals.

```{r}
ggplot(titanic, aes(age,survived)) + 
  + geom_jitter(height=0.05, width=0, alpha=0.5)  + geom_smooth(method="loess")
```

## Exercise 9
Create separate loess curves for males and females. Put the legend above the graph.


```{r}
ggplot(titanic, aes(age,survived,colour=sex)) + 
  geom_jitter(height=0.05, width=0,alpha=0.5) + geom_smooth(method="loess") + 
  theme(legend.position='top')
```


```{r}
# Extra: we change the label and the colour palette
ggplot(titanic, aes(age,survived,colour=sex)) + 
  geom_jitter(height=0.05,alpha=0.5) + geom_smooth(method="loess") +
  scale_colour_brewer(name="gender",label=c("F","M"),palette='Set1') + 
  theme(legend.position='top')
```

## Exercise 10
Split the graph into three panels based on passenger class. Truncate the range of the y-axis. Do this by setting the y-axis limits to c(-0.1,1.1) and see what happens.

```{r}
ggplot(titanic, aes(age,survived,colour=sex)) + 
  geom_jitter(height=0.05,alpha=0.3) + geom_smooth(method="loess") +
  facet_wrap(~pclass) + ylim(c(-0.1,1.1)) +  theme(legend.position='top')
```


```{r}
# Truncating the output is prevented by setting the limits within coord_cartesian
ggplot(titanic, aes(age,survived,colour=sex)) + 
  geom_jitter(height=0.05,alpha=0.3) + geom_smooth(method="loess") +
  facet_wrap(~pclass) + coord_cartesian(ylim=c(-0.1,1.1))
```
